---
title: "Reijn spectra and plots"
format: html
---

## Setup

```{r}
library(magrittr)
library(ICAMS)
library(mSigPlot)
library(ggplot2)
library(cowplot)
library(patchwork)
library(glue)
library(Cairo)
library(data.table)

sigFfile = "DRUP01030028T"
sig4file = "CPCT02100089T"

source("../code/Generate_Koh89_Koh476_catalog_0121.R")
```

## Helper functions

```{r}
stdread = function(ff) {
  read.delim(ff, sep = '\t', row.names = 1)
}
stdwrite = function(obj, ff) {
  write.table(obj, ff, sep = '\t', row.names = TRUE, col.names = NA)
}


myplot83 = function(xx, plot_title) {
  plot_83(
    xx,
    plot_title = plot_title,
    base_size = 11,
    text_size = 4,
    count_label_size = 3.5
  )
}

myplot476 = function(xx, plot_title) {
  plot_476(
    xx,
    plot_title = plot_title,
    base_size = 10,
    label_size = 3,
    text_size = 2.5,
    label_threshold_denominator = 4
  )
}

myplot476right = function(xx, plot_title) {
  plot_476_right(
    xx,
    plot_title = plot_title,
    base_size = 10,
    label_size = 3,
    text_size = 2.5,
    label_threshold_denominator = 4
  )
}
```

## Read Annotated VCFs from Reijn data

There are multiple samples in each VCF, as indicated by column `Sample`.

```{r}
mouse = "../Manuscript_data/Reijn.mouse.vcfs.with.annotation.txt"
rpe1 = "../Manuscript_data/Reijn.rpe1.vcfs.with.annotation.txt"

mvcf = fread(mouse) # Mice

rvcf = fread(rpe1) # Cells
```

## Split Reijn cell data by mutated and wild-type

```{r}
rvcf %>% dplyr::filter(Sample %in% c("AKO1_4", "BKO1_4")) -> cell_mut
fwrite(cell_mut, "../Manuscript_data/Reijn_mut_cell_annot.vcf")

rvcf %>% dplyr::filter(!Sample %in% c("AKO1_4", "BKO1_4")) -> cell_wt
fwrite(cell_wt, "../Manuscript_data/Reijn_wt_cell_annot.vcf")
```

## Get Reijn summary spectra catalogs

```{r}
GenerateKoh476CatalogfromAnnotateVcf(mvcf, "Sample") %>%
  rowSums() %>%
  as.data.frame() -> mouse476

colnames(mouse476) <- "all_mouse_mutations_summed"

stdwrite(mouse476, "mutated_mice.tsv")

```

## Plot mice (476, 89, and 83 types)

```{r}
plot_mice = function() {
  p4_1 = myplot476(mouse476, plot_title = "Mutated mice")

  xm4 = GenerateKoh89CatalogfromAnnotateVcf(mvcf, "Sample")
  xm_all = as.data.frame(rowSums(xm4))
  p8_1 = plot_89(xm_all, base_size = 20, plot_title = "Mutated mice")

  colnames(mvcf)[1:2] <- c("CHROM", "POS")
  all_mice_83 = ICAMS::VCFsToIDCatalogs(
    list(all_mice = mvcf[, 1:8]),
    ref.genome = "mm10"
  )
  all_mice_83[[1]]["DEL:T:1:5+", ] <- 0
  all_mice_83[[1]]["INS:T:1:5+", ] <- 0
  mouse83 = myplot83(
    all_mice_83[[1]],
    plot_title = "Mutated mice, ins/del T in long polyT <- 0"
  )
  return(c(mouse476 = p4_1, mouse89 = p8_1, mouse83 = mouse83))
}

mice = plot_mice()
```

## Plot cells (476 type)

```{r}
plot_cells476 = function() {
  r4 = GenerateKoh476CatalogfromAnnotateVcf(rvcf, "Sample")
  r4_mut = r4[, 1:2]
  stdwrite(r4_mut, "mutated_cells.tsv")
  r4_wt = r4[, 3:5]
  stdwrite(r4_mut, "wild_type_cells.tsv")
  r_all_mut = as.data.frame(rowSums(r4_mut))
  p4_2 = myplot476(r_all_mut, plot_title = "Mutated cells")
  r_all_wt = as.data.frame(rowSums(r4_wt))
  p4_3 = myplot476(r_all_wt, plot_title = "Wild-type cells")
  return(c(cell476_mut = p4_2, cell476_wt = p4_3))
}
cells476 = plot_cells476()
```

## Plot cells (89 type)

```{r}
plot_cells89 = function() {
  xr4 = GenerateKoh89CatalogfromAnnotateVcf(rvcf, "Sample")
  xr4_mut = xr4[, 1:2]
  xr4_wt = xr4[, 3:5]
  xr_all_mut = as.data.frame(rowSums(xr4_mut))
  p8_2 = plot_89(xr_all_mut, base_size = 20, plot_title = "Mutated cells")
  xr_all_wt = as.data.frame(rowSums(xr4_wt))
  p8_3 = plot_89(xr_all_wt, base_size = 20, plot_title = "Wild-type cells")
  return(c(cell89_mut = p8_2, cell89_wt = p8_3))
}
cells89 = plot_cells89()
```

## Plot cells (83 type)

```{r}
plot_cells83 = function() {
  rvcf %>%
    dplyr::select(CHROM, POS, ALT, REF, Sample) %>%
    dplyr::group_by(Sample) %>%
    dplyr::group_split() %>%
    ICAMS::VCFsToIDCatalogs(ref.genome = "hg38") %>%
    `$`(catalog) -> cats
  mut = cats[, 1:2]
  wt = cats[, 3:5]
  # browser()
  all_mut = as.data.frame(rowSums(mut))
  all_mut["DEL:T:1:5+", ] <- 0
  all_mut["INS:T:1:5+", ] <- 0
  cell83_mut = myplot83(
    all_mut,
    plot_title = "Mutated cells, ins/del T in long polyT <- 0"
  )
  all_wt = as.data.frame(rowSums(wt))
  all_wt["DEL:T:1:5+", ] <- 0
  all_wt["INS:T:1:5+", ] <- 0
  cell83_wt = myplot83(
    all_wt,
    plot_title = "Wild-type cells, ins/del T in long polyT <- 0"
  )
  return(c(cell83_mut, cell83_wt))
}
cells83 = plot_cells83()
```

## Signature plots (83 type)

```{r}
stdread("../Manuscript_data/Liu_et_al_final_83_type_signatures.tsv") %>%
  `[`("ID_F") %>%
  myplot83("ID_F") -> sig83F

stdread("../Manuscript_data/Liu_et_al_final_83_type_signatures.tsv") %>%
  `[`("C_ID4") %>%
  myplot83("C_ID4") -> sig834
```

## Signature plots (476 type)

```{r}
stdread("../Manuscript_data/Liu_et_al_final_476_type_signatures.tsv") %>%
  dplyr::select(dplyr::contains("InsDel_F")) -> sigF_catalog
sigF_catalog["Del2:M1", ]
# 0.2399
sigF_catalog["Del2:U2:R2", ]
# 0.2348

sigF_catalog %>%
  myplot476("InsDelF") -> sigF476

stdread("../Manuscript_data/Liu_et_al_final_476_type_signatures.tsv") %>%
  `[`("InsDel4") -> sig4_catalog
sig4_catalog["Del2:M1", ]
# 0.04781
sig4_catalog["Del2:U2:R2", ]
# 0.07488

sig4_catalog %>%
  myplot476("InsDel4") -> sig4476
```

## Linking tumor spectra

```{r}
stdread("../Manuscript_data/Liu_et_al_476_type_spectra.tsv") %>%
  dplyr::select(dplyr::contains(sigFfile)) -> sFlink_catalog

sFlink_catalog %>%
  myplot476(glue("InsDelF linking tumor {sigFfile}")) -> fF476
sFlink_catalog["Del2:M1", ]

# 339

sFlink_catalog["Del2:M1", ] / colSums(sFlink_catalog)

sFlink_catalog["Del2:U2:R2", ]
# 294

(sFlink_catalog["Del2:M1", ] + sFlink_catalog["Del2:U2:R2", ])

sFlink_catalog %>%
  filter(!grepl("\\([CT]\\)", rownames(sFlink_catalog))) %>%
  colSums()

stdread("../Manuscript_data/Liu_et_al_476_type_spectra.tsv") %>%
  dplyr::select(dplyr::contains(sig4file)) -> s4link_catalog
s4link_catalog["Del2:M1", ]
# 38
s4link_catalog["Del2:U2:R2", ]
# 77

s4link_catalog %>%
  myplot476(glue("InsDel4 linking tumor {sig4file}")) -> f4476
```

## Combined 476-type plot

```{r}
pw476 <- (sigF476 /
  fF476 /
  mice[["mouse476"]] /
  cells476[["cell476_mut"]] /
  cells476[["cell476_wt"]] /
  sig4476 /
  f4476) &
  theme(
    # Removes the "Indel Type" text and its occupied space
    axis.title.x = element_blank(),
    # Removes the x-axis text (R11A, etc.) if you don't need them for every plot
    # axis.text.x = element_blank(),
    plot.margin = margin(t = 0, r = 0, b = -0.2, l = 0, unit = "cm")
  ) &
  coord_cartesian(clip = "on") &
  scale_y_continuous(expand = c(0, 0))

pa <- plot_annotation(
  caption = "Indel Type",
  theme = theme(
    # Centers the caption at the bottom of the page
    plot.caption = element_text(hjust = 0.5, size = 12, face = "plain"),
    # This remains your 0.75-inch page border
    plot.margin = margin(0.75, 0.75, 0.75, 0.75, "in")
  )
)

ggsave(
  "F476.pdf",
  pw476 + pa,
  device = cairo_pdf,
  width = 8.5,
  height = 11,
  units = "in"
)
```

## Combined 476-type plot (right)

```{r}
#| label: plot476right

plot_mice_right = function() {
  myplot476right(mouse476, plot_title = "Mutated mice")
}

plot_cells476_right = function() {
  r4 = GenerateKoh476CatalogfromAnnotateVcf(rvcf, "Sample")
  r4_mut = r4[, 1:2]
  r4_wt = r4[, 3:5]
  r_all_mut = as.data.frame(rowSums(r4_mut))
  p4_2 = myplot476right(r_all_mut, plot_title = "Mutated cells")
  r_all_wt = as.data.frame(rowSums(r4_wt))
  p4_3 = myplot476right(r_all_wt, plot_title = "Wild-type cells")
  return(c(cell476_mut = p4_2, cell476_wt = p4_3))
}

mice_r = list(mouse476 = plot_mice_right())
cells476_r = plot_cells476_right()

sigF476_r = sigF_catalog %>% myplot476right("InsDelF")
sig4476_r = sig4_catalog %>% myplot476right("InsDel4")
fF476_r = sFlink_catalog %>%
  myplot476right(glue("InsDelF linking tumor {sigFfile}"))
f4476_r = s4link_catalog %>%
  myplot476right(glue("InsDel4 linking tumor {sig4file}"))

pw476r <- (sigF476_r /
  fF476_r /
  mice_r[["mouse476"]] /
  cells476_r[["cell476_mut"]] /
  cells476_r[["cell476_wt"]] /
  sig4476_r /
  f4476_r) &
  theme(
    axis.title.x = element_blank(),
    plot.margin = margin(t = 0, r = 0, b = -0.2, l = 0, unit = "cm")
  ) &
  coord_cartesian(clip = "on") &
  scale_y_continuous(expand = c(0, 0))

ggsave(
  "F476_right.pdf",
  pw476r + pa,
  device = cairo_pdf,
  width = 8.5,
  height = 11,
  units = "in"
)
```

## Combined 89-type plot

```{r}
pw89 <- (mice[["mouse89"]] /
  cells89[["cell89_mut"]] /
  cells89[["cell89_wt"]]) &
  theme(
    # Removes the "Indel Type" text and its occupied space
    axis.title.x = element_blank(),
    # Removes the x-axis text (R11A, etc.) if you don't need them for every plot
    # axis.text.x = element_blank(),
    plot.margin = margin(t = 0, r = 0, b = -0.2, l = 0, unit = "cm")
  ) &
  coord_cartesian(clip = "on") &
  scale_y_continuous(expand = c(0, 0))

ggsave(
  "F89.pdf",
  pw89 + pa,
  device = cairo_pdf,
  width = 8.5,
  height = 11,
  units = "in"
)
```

## Combined 83-type plot

```{r}
pw83 <- (sig83F /
  mice[["mouse83"]] /
  cells83[[1]] /
  cells83[[2]] /
  sig834) &
  theme(
    # Removes the "Indel Type" text and its occupied space
    axis.title.x = element_blank(),
    # Removes the x-axis text (R11A, etc.) if you don't need them for every plot
    # axis.text.x = element_blank(),
    plot.margin = margin(t = 0.5, r = 0, b = 0.5, l = 0, unit = "cm")
  ) &
  coord_cartesian(clip = "off") &
  scale_y_continuous(expand = c(0, 0))


ggsave(
  "F83.pdf",
  pw83 + pa,
  device = cairo_pdf,
  width = 10,
  height = 15,
  units = "in"
)
```
