---
title: "Interactive Dendrogram v2: Seed-filtered Signatures"
format: html
server: shiny
date: now
date-format: "YYYY-MM-DD HH:mm"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| context: setup

library(shiny)
library(plotly)
library(ggdendro)
library(lsa)
library(mSigPlot)
library(gridExtra)
library(ggplot2)

source("dendro2_helpers.R")

# Pre-load both full datasets
all_data <- list()
for (ds in c("Koh89", "Koh476")) {
  all_data[[ds]] <- load_all_signatures(ds)
}

# Liu signature names for each dataset type
liu_sig_names <- list(
  Koh89  = colnames(all_data[["Koh89"]]$combined)[
    all_data[["Koh89"]]$source_vec == "Liu"],
  Koh476 = colnames(all_data[["Koh476"]]$combined)[
    all_data[["Koh476"]]$source_vec == "Liu"]
)
```

```{r}
#| label: wide-layout
tags$style(".container-fluid { max-width: 100% !important; padding: 0 15px; }")
```

This interactive dendrogram loads **all** CAP9/NoCAP signature extractions and lets you
pick a seed Liu signature. Only signatures with cosine similarity >= threshold to
the seed are shown in the dendrogram.

## Controls

```{r}
#| label: ui-controls
#| panel: input

selectInput(
  "dataset", "Dataset:",
  choices = c("Koh89" = "Koh89", "Koh476" = "Koh476"),
  selected = "Koh89"
)

selectInput(
  "seed_signature", "Seed (Liu) Signature:",
  choices = liu_sig_names[["Koh89"]]
)

numericInput(
  "cosine_sim_to_seed", "Min Cosine Similarity to Seed:",
  value = 0.8, min = 0, max = 1, step = 0.05
)

actionButton("update_dendro", "Update Dendrogram")

uiOutput("selection_display")
```

```{r}
#| label: server-logic
#| context: server

# Update seed dropdown when dataset changes
observeEvent(input$dataset, {
  updateSelectInput(
    session, "seed_signature",
    choices = liu_sig_names[[input$dataset]]
  )
})

# Reactive: selected signatures (max 10)
selected_sigs <- reactiveVal(character(0))

# Clear selection on dataset switch or dendro update
observeEvent(input$dataset, { selected_sigs(character(0)) })

# Filtered + clustered data, recomputed only on button click
filtered_result <- eventReactive(input$update_dendro, {
  ds <- isolate(input$dataset)
  seed <- isolate(input$seed_signature)
  threshold <- isolate(input$cosine_sim_to_seed)

  combined <- all_data[[ds]]$combined
  source_vec <- all_data[[ds]]$source_vec

  filtered <- filter_by_seed(combined, seed, threshold)
  filtered_source <- source_vec[colnames(filtered)]

  dendro <- compute_dendrogram(filtered)

  selected_sigs(character(0))

  list(
    combined = filtered,
    source_vec = filtered_source,
    dendro = dendro
  )
}, ignoreNULL = FALSE)

# Render selection badges
output$selection_display <- renderUI({
  sigs <- selected_sigs()
  if (length(sigs) == 0) {
    return(tags$p(
      tags$em("Click leaves in the dendrogram to select signatures (max 10).")
    ))
  }

  res <- filtered_result()
  sv <- res$source_vec

  badges <- lapply(sigs, function(s) {
    col <- dendro2_colors[sv[s]]
    tags$span(
      s,
      style = paste0(
        "display: inline-block; margin: 2px 4px; padding: 3px 8px; ",
        "border-radius: 4px; color: white; font-size: 0.85em; ",
        "background-color: ", col, ";"
      )
    )
  })

  tagList(
    tags$p(tags$strong(paste0("Selected (", length(sigs), "/10): ")),
           badges),
    actionButton("clear_selection", "Clear All",
                 style = "margin-top: 4px;")
  )
})

observeEvent(input$clear_selection, {
  selected_sigs(character(0))
})
```

## Dendrogram

```{r}
#| label: dendrogram-plot

plotlyOutput("dendrogram", height = "400px", width = "100%")
```

```{r}
#| label: dendrogram-server
#| context: server

output$dendrogram <- renderPlotly({
  res <- filtered_result()
  build_plotly_dendrogram(
    res$dendro$dend_data,
    res$source_vec,
    colors = dendro2_colors
  )
})

# Handle click to toggle signature selection
observeEvent(event_data("plotly_click", source = "dendrogram"), {
  click <- event_data("plotly_click", source = "dendrogram")
  if (!is.null(click)) {
    sig_name <- click$customdata
    if (is.list(sig_name)) sig_name <- sig_name[[1]]
    if (!is.null(sig_name) && nzchar(sig_name)) {
      current <- selected_sigs()
      if (sig_name %in% current) {
        selected_sigs(setdiff(current, sig_name))
      } else if (length(current) < 10) {
        selected_sigs(c(current, sig_name))
      }
    }
  }
})
```

## Signature Profiles

```{r}
#| label: sig-profiles

uiOutput("sig_plots_ui")
```

```{r}
#| label: sig-profiles-server
#| context: server

output$sig_plots_ui <- renderUI({
  sigs <- selected_sigs()
  if (length(sigs) == 0) {
    return(tags$p(tags$em("Select signatures above to view their profiles.")))
  }
  height_per <- if (input$dataset == "Koh89") 300 else 400
  total_height <- paste0(length(sigs) * height_per, "px")
  plotOutput("sig_plots", height = total_height, width = "100%")
})

sig_plots_height <- reactive({
  sigs <- selected_sigs()
  if (length(sigs) == 0) return(300)
  height_per <- if (input$dataset == "Koh89") 300 else 400
  length(sigs) * height_per
})

output$sig_plots <- renderPlot({
  sigs <- selected_sigs()
  req(length(sigs) > 0)

  res <- filtered_result()
  combined <- res$combined
  plot_fn <- if (input$dataset == "Koh89") mSigPlot::plot_89 else mSigPlot::plot_476

  plots <- lapply(sigs, function(s) {
    sig_data <- combined[, s, drop = FALSE]
    plot_fn(sig_data, plot_title = s)
  })

  gridExtra::grid.arrange(grobs = plots, ncol = 1)
}, res = 100,
   width = function() { if (input$dataset == "Koh476") 1400 else 900 },
   height = function() { sig_plots_height() })
```

## Cosine Similarity

```{r}
#| label: cosine-heatmap

plotlyOutput("cosine_heatmap", height = "500px")
```

```{r}
#| label: cosine-heatmap-server
#| context: server

output$cosine_heatmap <- renderPlotly({
  sigs <- selected_sigs()
  if (length(sigs) < 2) {
    return(
      plot_ly(type = "scatter", mode = "markers") |>
        layout(
          title = "Select at least 2 signatures to see cosine similarity.",
          xaxis = list(visible = FALSE),
          yaxis = list(visible = FALSE)
        )
    )
  }

  res <- filtered_result()
  cos_mat <- res$dendro$cosine_sim[sigs, sigs]

  text_mat <- matrix(
    sprintf("%.3f", cos_mat),
    nrow = nrow(cos_mat)
  )

  plot_ly(
    x = sigs, y = sigs,
    z = cos_mat,
    text = text_mat,
    type = "heatmap",
    colorscale = list(c(0, "#FFFFFF"), c(1, "#FF0000")),
    zmin = 0, zmax = 1,
    hovertemplate = "%{y} vs %{x}<br>Cosine: %{text}<extra></extra>"
  ) |>
    layout(
      title = "Pairwise Cosine Similarity",
      xaxis = list(title = "", tickangle = -45),
      yaxis = list(title = "", autorange = "reversed"),
      margin = list(l = 120, b = 120)
    )
})
```

## About

- **Source colors**: Liu (red), C.H = CAP9 Hartwig (green), C.P = CAP9 PCAWG (purple), C.PH = CAP9 PCAWG+Hartwig (orange), N.H = NoCAP Hartwig (blue)
- **Workflow**: Select a dataset, pick a seed Liu signature, set cosine similarity threshold, click "Update Dendrogram"
- **Click** a leaf marker to select/deselect it (shown in badges above)
- **Zoom** into crowded regions using plotly's built-in zoom
- Signature profiles are rendered with mSigPlot (`plot_89` or `plot_476`)
- Max 10 signatures can be selected at once
