---
title: "Interactive Signature Assignments"
format: html
server: shiny
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| context: setup

library(shiny)
library(plotly)
library(jsonlite)
library(mSigPlot)
library(gridExtra)

# Load helper functions
source("hamburger_helpers.R")

# Data directory
data_dir <- "../Manuscript_data/"

# Load assignment matrices
assignments_83_89 <- read.delim(

  file.path(data_dir, "Liu_et_al_83_plus_89_as_83_type_signature_assignments.tsv"),
  row.names = 1,
  check.names = FALSE
)

assignments_648 <- read.delim(
  file.path(data_dir, "Liu_et_al_648_type_signature_assignments.tsv"),
  row.names = 1,
  check.names = FALSE
)

# Load spectra files
spectra_83 <- read.delim(
  file.path(data_dir, "Liu_et_al_83_type_spectra.tsv"),
  row.names = 1,
  check.names = FALSE
)

spectra_89 <- read.delim(
  file.path(data_dir, "Liu_et_al_89_type_spectra.tsv"),
  row.names = 1,
  check.names = FALSE
)

spectra_476 <- read.delim(
  file.path(data_dir, "Liu_et_al_476_type_spectra.tsv"),
  row.names = 1,
  check.names = FALSE
)

# Load signature files
sigs_83 <- read.delim(
  file.path(data_dir, "Liu_et_al_final_83_type_signatures.tsv"),
  row.names = 1,
  check.names = FALSE
)

sigs_89 <- read.delim(
  file.path(data_dir, "Liu_et_al_final_89_type_signatures.tsv"),
  row.names = 1,
  check.names = FALSE
)

sigs_476 <- read.delim(
  file.path(data_dir, "Liu_et_al_final_476_type_signatures.tsv"),
  row.names = 1,
  check.names = FALSE
)

# Load signature mapping (89-type to 83-type correspondence)
sig_mapping <- read.delim(
  file.path(data_dir, "89type_to_83type_connection.tsv"),
  check.names = FALSE
)

# Get signature lists
signatures_83_89 <- rownames(assignments_83_89)
signatures_648 <- rownames(assignments_648)
```

This interactive visualization shows the distribution of signature mutations across cancer types ("hamburger plots"). Click on any point to view that tumor's mutational spectrum in ID83, ID89, and ID476 classifications using mSigPlot.

## Select Assignment Type and Signature

```{r}
#| label: ui-controls
#| panel: input

selectInput(
 "assignment_type",
 "Assignment Type:",
 choices = c(
    "83+89 as 83-type (33 signatures)" = "83_89",
    "648-type (44 signatures)" = "648"
  ),
  selected = "83_89"
)

uiOutput("signature_selector")
```

```{r}
#| label: signature-selector-server
#| context: server

output$signature_selector <- renderUI({
  sigs <- if (input$assignment_type == "83_89") signatures_83_89 else signatures_648
  selectInput("signature", "Signature:", choices = sigs)
})
```

## Hamburger Plot

Click on any point to select a tumor and view its spectrum below.

```{r}
#| label: hamburger-plot

plotlyOutput("hamburger_plot", height = "500px")
```

```{r}
#| label: hamburger-plot-server
#| context: server

# Reactive for current hamburger data
hamburger_data <- reactive({
  req(input$signature)

  assignment_matrix <- if (input$assignment_type == "83_89") {
    assignments_83_89
  } else {
    assignments_648
  }

  get_hamburger_data(input$signature, assignment_matrix, min_samples = 3)
})

# Render hamburger plot
output$hamburger_plot <- renderPlotly({
  df <- hamburger_data()

  if (is.null(df) || nrow(df) == 0) {
    return(plotly_empty() |> layout(title = "No data available for this signature"))
  }

  type_order <- attr(df, "type_order")
  medians <- attr(df, "medians")
  n_types <- length(type_order)

  bg_colors <- c("#EDF8B1", "#2C7FB8")

  # Create alternating background shapes
  bg_shapes <- lapply(seq_len(n_types), function(i) {
    list(
      type = "rect",
      x0 = i - 0.5,
      x1 = i + 0.5,
      y0 = 0,
      y1 = 1,
      xref = "x",
      yref = "paper",
      fillcolor = bg_colors[(i - 1) %% 2 + 1],
      opacity = 0.2,
      line = list(width = 0),
      layer = "below"
    )
  })

  # Create median line shapes
  median_shapes <- lapply(seq_len(nrow(medians)), function(i) {
    x_pos <- which(type_order == medians$cancer_type[i])
    list(
      type = "line",
      x0 = x_pos - 0.4,
      x1 = x_pos + 0.4,
      y0 = medians$median_mutations[i],
      y1 = medians$median_mutations[i],
      xref = "x",
      yref = "y",
      line = list(color = "red", width = 2, dash = "dash"),
      layer = "above"
    )
  })

  all_shapes <- c(bg_shapes, median_shapes)

  # Create hover text
  df$hover_text <- paste0(
    "<b>Sample:</b> ", df$sample_id, "<br>",
    "<b>Cancer Type:</b> ", df$cancer_type, "<br>",
    "<b>Mutations:</b> ", round(df$mutations, 1)
  )

  plot_ly(
    data = df,
    x = ~x_plot,
    y = ~mutations,
    type = "scatter",
    mode = "markers",
    marker = list(size = 6, color = "black", opacity = 0.6),
    text = ~hover_text,
    hoverinfo = "text",
    customdata = ~sample_id,
    source = "hamburger"
  ) |>
    layout(
      title = list(text = paste0(input$signature, " mutations"), x = 0.5),
      xaxis = list(
        title = "",
        tickmode = "array",
        tickvals = seq_len(n_types),
        ticktext = as.character(type_order),
        tickangle = 60,
        side = "top",
        showgrid = FALSE
      ),
      yaxis = list(
        title = "Number of Mutations",
        type = "log",
        showgrid = TRUE
      ),
      shapes = all_shapes,
      hovermode = "closest",
      margin = list(t = 120, b = 50, l = 60, r = 20)
    ) |>
    config(displayModeBar = TRUE, displaylogo = FALSE) |>
    event_register("plotly_click")
})

# Track selected sample from click
selected_sample <- reactiveVal(NULL)

observeEvent(event_data("plotly_click", source = "hamburger"), {
  click_data <- event_data("plotly_click", source = "hamburger")
  if (!is.null(click_data)) {
    selected_sample(click_data$customdata[[1]])
  }
})
```

## Selected Tumor Spectrum

```{r}
#| label: selected-info

uiOutput("selected_info")
```

```{r}
#| label: selected-info-server
#| context: server

output$selected_info <- renderUI({
  sample_id <- selected_sample()
  if (is.null(sample_id)) {
    tags$p(tags$strong("Select a point from the hamburger plot above to view spectrum."))
  } else {
    tags$p(tags$strong("Selected Tumor: "), sample_id)
  }
})
```

::: {.panel-tabset}

### 83-type

```{r}
#| label: spectrum-83

plotOutput("spectrum_83", height = "500px")
```

```{r}
#| label: spectrum-83-server
#| context: server

output$spectrum_83 <- renderPlot({
  sample_id <- selected_sample()
  sig_name <- input$signature
  req(sample_id, sig_name)

  if (!sample_id %in% colnames(spectra_83)) {
    plot.new()
    text(0.5, 0.5, paste("No ID83 data for", sample_id))
    return()
  }

  # Get the 83-type signature name
  if (input$assignment_type == "83_89") {
    sig_83_name <- sig_name  # Already 83-type
  } else {
    # Map from 89-type to 83-type
    idx <- match(sig_name, sig_mapping$InDel89)
    sig_83_name <- if (!is.na(idx)) sig_mapping$InDel83[idx] else NULL
  }

  spectrum <- spectra_83[, sample_id, drop = FALSE]
  p1 <- mSigPlot::plot_83(spectrum, plot_title = paste("Spectrum:", sample_id))

  if (!is.null(sig_83_name) && sig_83_name %in% colnames(sigs_83)) {
    sig_data <- sigs_83[, sig_83_name, drop = FALSE]
    p2 <- mSigPlot::plot_83(sig_data, plot_title = paste("Signature:", sig_83_name))
    gridExtra::grid.arrange(p1, p2, ncol = 1)
  } else {
    p1
  }
}, res = 100)
```

### 89-type

```{r}
#| label: spectrum-89

plotOutput("spectrum_89", height = "500px")
```

```{r}
#| label: spectrum-89-server
#| context: server

output$spectrum_89 <- renderPlot({
  sample_id <- selected_sample()
  sig_name <- input$signature
  req(sample_id, sig_name)

  if (!sample_id %in% colnames(spectra_89)) {
    plot.new()
    text(0.5, 0.5, paste("No ID89 data for", sample_id))
    return()
  }

  # Get the 89-type signature name
  if (input$assignment_type == "648") {
    sig_89_name <- sig_name  # Already 89-type
  } else {
    # Map from 83-type to 89-type (find first matching 89-type)
    idx <- which(sig_mapping$InDel83 == sig_name)
    sig_89_name <- if (length(idx) > 0) sig_mapping$InDel89[idx[1]] else NULL
  }

  spectrum <- spectra_89[, sample_id, drop = FALSE]
  p1 <- mSigPlot::plot_89(spectrum, plot_title = paste("Spectrum:", sample_id))

  if (!is.null(sig_89_name) && sig_89_name %in% colnames(sigs_89)) {
    sig_data <- sigs_89[, sig_89_name, drop = FALSE]
    p2 <- mSigPlot::plot_89(sig_data, plot_title = paste("Signature:", sig_89_name))
    gridExtra::grid.arrange(p1, p2, ncol = 1)
  } else {
    p1
  }
}, res = 100)
```

### 476-type

```{r}
#| label: spectrum-476

plotOutput("spectrum_476", height = "600px", width = "100%")
```

```{r}
#| label: spectrum-476-server
#| context: server

output$spectrum_476 <- renderPlot({
  sample_id <- selected_sample()
  sig_name <- input$signature
  req(sample_id, sig_name)

  if (!sample_id %in% colnames(spectra_476)) {
    plot.new()
    text(0.5, 0.5, paste("No ID476 data for", sample_id))
    return()
  }

  # Get the 476-type signature name (uses same names as 89-type)
  if (input$assignment_type == "648") {
    # Already 89-type name, use directly
    sig_476_name <- sig_name
  } else {
    # Map from 83-type to 89-type (find first matching)
    idx <- which(sig_mapping$InDel83 == sig_name)
    sig_476_name <- if (length(idx) > 0) sig_mapping$InDel89[idx[1]] else NULL
  }

  spectrum <- spectra_476[, sample_id, drop = FALSE]
  p1 <- mSigPlot::plot_476(spectrum, plot_title = paste("Spectrum:", sample_id))

  if (!is.null(sig_476_name) && sig_476_name %in% colnames(sigs_476)) {
    sig_data <- sigs_476[, sig_476_name, drop = FALSE]
    p2 <- mSigPlot::plot_476(sig_data, plot_title = paste("Signature:", sig_476_name))
    gridExtra::grid.arrange(p1, p2, ncol = 1)
  } else {
    p1
  }
}, res = 100, width = 1400, height = 600)
```

:::

## About This Visualization

This page displays **hamburger plots** (also known as snake plots) showing how mutations attributed to each signature are distributed across cancer types.

### How to Use

1. **Select an assignment type**: Choose between the 83+89 as 83-type assignments (33 signatures) or the 648-type assignments (44 signatures).

2. **Select a signature**: Pick a signature from the dropdown to view its hamburger plot.

3. **Click on a point**: Each point represents one tumor. Click on a point to view that tumor's full mutational spectrum rendered with mSigPlot in ID83, ID89, and ID476 classifications.

### Reading the Hamburger Plot

- **X-axis**: Cancer types, ordered by median mutation count (ascending)
- **Y-axis**: Number of mutations attributed to this signature (log scale)
- **Points**: Individual tumors, spread horizontally within each cancer type column
- **Red dashed lines**: Median mutation count for each cancer type
- **Alternating backgrounds**: Help distinguish adjacent cancer type columns
