---
title: "Clustering CAP9 Koh89 Signatures (min similarity = `r params$min_similarity`)"
author: "Steve Rozen (sr110@duke.edu, steverozen@pm.me)"
date: now
date-format: "YYYY-MM-DD HH:mm"
params:
  min_similarity: 0.95
format:
  html:
    page-layout: full
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    include-in-header:
      text: |
        <style>
          .quarto-container, .page-columns, .main-container {
            max-width: 100% !important;
          }
        </style>
---

```{r setup}
#| message: false
library(here)
library(ggplot2)
library(ggdendro)
library(cluster)
library(mSigPlot)
library(DT)
library(plotly)
library(reshape2)

source(here::here("code", "dendro2_helpers.R"))
```

## Parameters

```{r params}
min_similarity   <- params$min_similarity
min_cluster_size <- 3
sig_dir <- here::here("Manuscript_data", "Mo_CAP9_analysis", "Signatures", "Koh89")
catalog_dir <- here::here("Manuscript_data", "Mo_CAP9_analysis", "Catalogs")
```

## Load and combine signatures

```{r load-sigs}
sig_files <- Sys.glob(file.path(sig_dir, "CAP9.*.txt"))
cat("Found", length(sig_files), "CAP9 files\n")

all_sigs <- lapply(sig_files, rename_sigs2)

# Filter to expected row count (89)
nrows <- sapply(all_sigs, nrow)
all_sigs <- all_sigs[nrows == 89]
cat("Files with 89 rows:", length(all_sigs), "\n")

# Assign ICAMS canonical row names
cat_files <- Sys.glob(file.path(catalog_dir, "*Koh89*.txt"))
cat_rownames <- rownames(read.table(
  cat_files[1], header = TRUE, sep = "\t",
  row.names = 1, check.names = FALSE
))
stopifnot(identical(cat_rownames, ICAMS::catalog.row.order$ID89))

for (i in seq_along(all_sigs)) {
  rownames(all_sigs[[i]]) <- cat_rownames
}

# Build source vector
source_vec <- character(0)
for (s in all_sigs) {
  grp <- attr(s, "source_group")
  source_vec <- c(source_vec, setNames(rep(grp, ncol(s)), colnames(s)))
}

# Combine into one matrix
combined <- as.matrix(do.call(cbind, all_sigs))
cat("Combined matrix:", nrow(combined), "rows x", ncol(combined), "columns\n")
```

## Cosine similarity matrix

```{r cosine-sim}
combined_t    <- t(combined)
combined_norm <- combined_t / sqrt(rowSums(combined_t^2))
cosine_sim    <- combined_norm %*% t(combined_norm)
cosine_dist   <- as.dist(1 - cosine_sim)
```

## Hierarchical clustering (complete linkage)

Complete linkage guarantees that all pairs within a cluster have cosine
similarity $\geq$ `r min_similarity` when we cut at height
`r 1 - min_similarity`.

```{r clustering}
hc <- hclust(cosine_dist, method = "complete")
cut_height <- 1 - min_similarity
clusters <- cutree(hc, h = cut_height)
n_clusters <- length(unique(clusters))
cat("Number of clusters at height", cut_height, ":", n_clusters, "\n")
```

## Compute medoids and filter clusters

```{r medoids}
compute_medoid <- function(members, cos_sim_mat) {
  if (length(members) == 1) return(members)
  sub_sim <- cos_sim_mat[members, members, drop = FALSE]
  diag(sub_sim) <- NA
  avg_sim <- rowMeans(sub_sim, na.rm = TRUE)
  names(which.max(avg_sim))
}

cluster_ids <- sort(unique(clusters))
summary_list <- list()
medoid_names <- character(0)

for (cid in cluster_ids) {
  members <- names(clusters[clusters == cid])
  size <- length(members)

  if (size == 1) {
    min_cs <- mean_cs <- max_cs <- NA
  } else {
    sub_sim <- cosine_sim[members, members, drop = FALSE]
    pairwise <- sub_sim[lower.tri(sub_sim)]
    min_cs  <- min(pairwise)
    mean_cs <- mean(pairwise)
    max_cs  <- max(pairwise)
  }

  medoid <- compute_medoid(members, cosine_sim)
  retained <- size >= min_cluster_size

  summary_list[[length(summary_list) + 1]] <- data.frame(
    cluster_id   = cid,
    size         = size,
    retained     = retained,
    medoid       = medoid,
    min_cos_sim  = round(min_cs, 4),
    mean_cos_sim = round(mean_cs, 4),
    max_cos_sim  = round(max_cs, 4),
    members      = paste(members, collapse = "; "),
    stringsAsFactors = FALSE
  )

  if (retained) {
    medoid_names <- c(medoid_names, setNames(cid, medoid))
  }
}

summary_df <- do.call(rbind, summary_list)
retained_df <- summary_df[summary_df$retained, ]
dropped_df  <- summary_df[!summary_df$retained, ]
```

## Summary statistics

```{r summary-stats}
#| results: asis
cat("- **Total signatures loaded:**", ncol(combined), "\n")
cat("- **Total clusters:**", n_clusters, "\n")
cat("- **Retained clusters** (size >=", min_cluster_size, "):",
    nrow(retained_df), "\n")
cat("- **Dropped clusters** (size <", min_cluster_size, "):",
    nrow(dropped_df), "\n")
cat("- **Signatures in retained clusters:**",
    sum(retained_df$size), "\n")
cat("- **Signatures dropped:**",
    ncol(combined) - sum(retained_df$size), "\n")

if (any(!is.na(retained_df$min_cos_sim))) {
  cat("- **Global min within-cluster cosine similarity:**",
      min(retained_df$min_cos_sim, na.rm = TRUE), "\n")
}
```

## Cluster summary table

```{r summary-table}
display_df <- retained_df[, c("cluster_id", "size", "medoid",
                               "min_cos_sim", "mean_cos_sim", "max_cos_sim")]
DT::datatable(
  display_df,
  rownames = FALSE,
  options = list(pageLength = 25, scrollX = TRUE),
  caption = paste("Retained clusters (size >=", min_cluster_size, ")")
)
```

## Dendrogram

:::{.column-screen}
```{r dendrogram}
#| out-width: "100%"
dend_data <- dendro_data(hc)

# Compute x-range for cluster 1 to set initial zoom
cluster1_members <- names(clusters[clusters == 1])
cluster1_x <- dend_data$labels$x[dend_data$labels$label %in% cluster1_members]
cluster1_span <- max(cluster1_x) - min(cluster1_x)
center <- mean(range(cluster1_x))
half_width <- (cluster1_span / 2 + 1) * 12
initial_xrange <- c(center - half_width, center + half_width)

build_plotly_dendrogram(
  dend_data      = dend_data,
  source_vec     = source_vec,
  cut_height     = cut_height,
  medoids        = names(medoid_names),
  initial_xrange = initial_xrange,
  initial_yrange = c(-0.12, 0.3),
  clusters       = clusters,
  strip_prefix   = "C\\."
)
```
:::


## Medoid + reference dendrogram

Dendrogram of medoid signatures together with Liu et al. and Koh et al.
reference signatures. Check the box to include all CAP9 extraction signatures.

```{r medoid-ref-load}
#| message: false
# Load Liu reference signatures
liu_sigs <- read.table(
  here::here("Manuscript_data", "Liu_et_al_final_89_type_signatures.tsv"),
  header = TRUE, sep = "\t", row.names = 1, check.names = FALSE
)

# Load Koh reference signatures
koh_sigs <- read.table(
  here::here("Manuscript_data", "Koh_signatures.tsv"),
  header = TRUE, sep = "\t", row.names = 1, check.names = FALSE
)

liu_source <- setNames(rep("Liu", ncol(liu_sigs)), colnames(liu_sigs))
koh_source <- setNames(rep("Koh", ncol(koh_sigs)), colnames(koh_sigs))

# --- Medoids-only version ---
medoid_mat_ref <- combined[, names(medoid_names), drop = FALSE]
ref_combined_med <- cbind(medoid_mat_ref, as.matrix(liu_sigs), as.matrix(koh_sigs))
ref_source_med <- c(
  setNames(source_vec[names(medoid_names)], names(medoid_names)),
  liu_source, koh_source
)

# --- All-signatures version ---
ref_combined_all <- cbind(combined, as.matrix(liu_sigs), as.matrix(koh_sigs))
ref_source_all <- c(source_vec, liu_source, koh_source)
```

:::{.column-screen}

```{=html}
<div style="margin-bottom: 8px;">
  <label style="font-size: 1.1em; cursor: pointer;">
    <input type="checkbox" id="toggle-all-sigs"
           onchange="
             document.getElementById('dend-medoids-only').style.display =
               this.checked ? 'none' : '';
             document.getElementById('dend-all-sigs').style.display =
               this.checked ? '' : 'none';
             /* Trigger plotly relayout so the revealed plot sizes correctly */
             var el = document.getElementById('dend-all-sigs');
             if (this.checked && el.querySelector('.plotly')) {
               window.dispatchEvent(new Event('resize'));
             }
           ">
    Include all extraction signatures
  </label>
</div>
```

::: {#dend-medoids-only}
```{r medoid-ref-dendrogram-med}
#| out-width: "100%"
ref_dend_med <- compute_dendrogram(ref_combined_med)

build_plotly_dendrogram(
  dend_data      = ref_dend_med$dend_data,
  source_vec     = ref_source_med,
  cut_height     = cut_height,
  medoids        = names(medoid_names),
  initial_yrange = c(-0.12, 0.2),
  clusters       = clusters,
  strip_prefix   = "C\\."
) |> layout(height = 900)
```
:::

::: {#dend-all-sigs style="display:none;"}
```{r medoid-ref-dendrogram-all}
#| out-width: "100%"
ref_dend_all <- compute_dendrogram(ref_combined_all)

# Compute initial x-range for cluster 1 (same logic as first dendrogram)
c1_members_all <- names(clusters[clusters == 1])
c1_x_all <- ref_dend_all$dend_data$labels$x[
  ref_dend_all$dend_data$labels$label %in% c1_members_all
]
c1_span_all <- max(c1_x_all) - min(c1_x_all)
c1_center_all <- mean(range(c1_x_all))
c1_half_all <- (c1_span_all / 2 + 1) * 12
initial_xrange_all <- c(c1_center_all - c1_half_all, c1_center_all + c1_half_all)

build_plotly_dendrogram(
  dend_data      = ref_dend_all$dend_data,
  source_vec     = ref_source_all,
  cut_height     = cut_height,
  medoids        = names(medoid_names),
  initial_xrange = initial_xrange_all,
  initial_yrange = c(-0.12, 0.2),
  clusters       = clusters,
  strip_prefix   = "C\\."
) |> layout(height = 900)
```
:::

:::

## Medoid pairwise cosine similarity heatmap

```{r medoid-heatmap}
#| fig-width: 12
#| fig-height: 11
cosine_cutoff <- 0.9

# Subset cosine similarity matrix to medoids only
medoid_sigs <- names(medoid_names)
med_sim <- cosine_sim[medoid_sigs, medoid_sigs]

# Strip "C." prefix from names to match dendrogram labels
short_names <- sub("^C\\.", "", medoid_sigs)
rownames(med_sim) <- short_names
colnames(med_sim) <- short_names

# Lower triangle only (NA for diagonal and upper)
med_sim_lower <- med_sim
med_sim_lower[upper.tri(med_sim_lower, diag = TRUE)] <- NA

melted <- reshape2::melt(med_sim_lower, na.rm = TRUE)
colnames(melted) <- c("Sig1", "Sig2", "Cosine")

melted$Sig1 <- factor(melted$Sig1, levels = short_names)
melted$Sig2 <- factor(melted$Sig2, levels = short_names)

# Label cells with cosine >= cutoff
melted$Label <- ifelse(
  melted$Cosine >= cosine_cutoff,
  sprintf("%.2f", melted$Cosine),
  ""
)

# Vertical line segments from high-similarity cells to x-axis
high_sim <- melted[melted$Cosine >= cosine_cutoff, ]
if (nrow(high_sim) > 0) {
  line_data <- data.frame(
    x    = as.numeric(high_sim$Sig2),
    xend = as.numeric(high_sim$Sig2),
    y    = as.numeric(high_sim$Sig1) - 0.5,
    yend = 0.5
  )
} else {
  line_data <- data.frame(
    x = numeric(), xend = numeric(),
    y = numeric(), yend = numeric()
  )
}

p_med_heat <- ggplot(melted, aes(x = Sig2, y = Sig1, fill = Cosine)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Label), size = 2.5) +
  scale_fill_gradient(
    low = "white", high = "red",
    limits = c(0, 1),
    name = "Cosine\nSimilarity"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
    axis.title  = element_blank(),
    panel.grid  = element_blank()
  ) +
  coord_fixed(clip = "off") +
  ggtitle("Medoid Pairwise Cosine Similarities (Lower Triangle)")

if (nrow(line_data) > 0) {
  p_med_heat <- p_med_heat +
    geom_segment(
      data = line_data,
      aes(x = x, xend = xend, y = y, yend = yend),
      inherit.aes = FALSE,
      color = "darkred", linetype = "solid", linewidth = 0.5
    )
}

print(p_med_heat)
```

## Cosine similarity heatmap

```{r heatmap}
#| fig-width: 12
#| fig-height: 11
# Order by dendrogram
dend_order <- hc$order
ordered_sim <- cosine_sim[dend_order, dend_order]

# Convert to long format for ggplot
n <- nrow(ordered_sim)
row_idx <- rep(seq_len(n), each = n)
col_idx <- rep(seq_len(n), times = n)
# Only lower triangle + diagonal for efficiency
keep <- col_idx <= row_idx
heatmap_df <- data.frame(
  x = col_idx[keep],
  y = row_idx[keep],
  sim = ordered_sim[cbind(row_idx[keep], col_idx[keep])]
)

p_heat <- ggplot(heatmap_df, aes(x = x, y = y, fill = sim)) +
  geom_raster() +
  scale_fill_gradient2(
    low = "white", mid = "orange", high = "red",
    midpoint = 0.9, limits = c(0.5, 1),
    name = "Cosine\nsimilarity"
  ) +
  coord_equal() +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(
    x = NULL, y = NULL,
    title = "Pairwise cosine similarity (dendrogram order)"
  )

print(p_heat)
```

## Silhouette analysis

Each signature's silhouette width measures how well it fits its assigned cluster
compared to its next-best alternative cluster. Values range from −1 to +1:

- **Near +1**: the signature is well-matched to its cluster and poorly matched to
  neighboring clusters.
- **Near 0**: the signature sits on the boundary between two clusters.
- **Negative**: the signature may be assigned to the wrong cluster.

Signatures are grouped by cluster and sorted by silhouette width within each
cluster. Alternating colors distinguish adjacent clusters.

```{r silhouette}
#| fig-width: 10
#| fig-height: 75
# Silhouette for retained clusters only
retained_members <- names(clusters[clusters %in% retained_df$cluster_id])
if (length(unique(clusters[retained_members])) > 1) {
  sil <- silhouette(clusters[retained_members], dist = as.dist(
    (1 - cosine_sim[retained_members, retained_members])
  ))
  sil_summary <- summary(sil)

  # Build data frame: sort by cluster, then by silhouette width within cluster
  sil_df <- data.frame(
    cluster = factor(sil[, "cluster"]),
    sil_width = sil[, "sil_width"]
  )
  sil_df <- sil_df[order(sil_df$cluster, -sil_df$sil_width), ]
  sil_df$y <- seq_len(nrow(sil_df))

  # Alternating colors per cluster
  clust_ids <- levels(sil_df$cluster)
  palette <- c("steelblue", "tomato")
  clust_color_map <- setNames(
    rep(palette, length.out = length(clust_ids)), clust_ids
  )
  sil_df$color <- clust_color_map[as.character(sil_df$cluster)]

  # Cluster label positions (midpoint of each cluster's y range)
  label_df <- do.call(rbind, lapply(clust_ids, function(cid) {
    rows <- sil_df[sil_df$cluster == cid, ]
    avg_sil <- mean(rows$sil_width)
    data.frame(
      cluster = cid,
      y = mean(range(rows$y)),
      label = sprintf("%s : %d | %.2f", cid, nrow(rows), avg_sil)
    )
  }))

  # Reverse y so cluster 1 is at the top (like base R silhouette)
  sil_df$y <- rev(sil_df$y)
  label_df$y <- sapply(clust_ids, function(cid) {
    mean(range(sil_df$y[sil_df$cluster == cid]))
  })

  p_sil <- ggplot(sil_df, aes(x = y, y = sil_width, fill = color)) +
    geom_col(width = 1, show.legend = FALSE) +
    scale_fill_identity() +
    scale_x_continuous(
      breaks = label_df$y,
      labels = label_df$label,
      expand = c(0.005, 0.005)
    ) +
    geom_hline(yintercept = sil_summary$avg.width, linetype = "dashed",
               color = "red") +
    coord_flip() +
    labs(
      y = "Silhouette width",
      x = NULL,
      title = "Silhouette plot (retained clusters)",
      subtitle = sprintf(
        "Average silhouette width: %.4f", sil_summary$avg.width
      )
    ) +
    theme_minimal() +
    theme(
      axis.text.y = element_text(size = 12, family = "mono"),
      panel.grid.major.y = element_blank(),
      panel.grid.minor.y = element_blank()
    )
  print(p_sil)
} else {
  cat("Only one retained cluster — silhouette not applicable.\n")
}
```

## Medoid signature plots

```{r load-catalogs}
#| message: false
# Load CAP9 catalogs for best-matching spectrum lookup
cap9_pcawg_cat <- as.matrix(read.table(
  here::here("Manuscript_data", "Mo_CAP9_analysis", "Catalogs",
             "CAP9.PCAWG.Koh89.catalog.txt"),
  header = TRUE, sep = "\t", row.names = 1, check.names = FALSE
))

cap9_hartwig_cat <- as.matrix(read.table(
  here::here("Manuscript_data", "Mo_CAP9_analysis", "Catalogs",
             "CAP9.Hartwig.Koh89.catalog.txt"),
  header = TRUE, sep = "\t", row.names = 1, check.names = FALSE
))

all_catalogs <- cbind(cap9_pcawg_cat, cap9_hartwig_cat)
cat("Combined catalog:", nrow(all_catalogs), "rows x",
    ncol(all_catalogs), "columns\n")
```

```{r medoid-plots}
#| fig-width: 10
#| fig-height: 4
#| results: asis
for (i in seq_len(nrow(retained_df))) {
  row <- retained_df[i, ]
  medoid_sig <- combined[, row$medoid, drop = FALSE]

  cat("\n\n### Cluster", row$cluster_id,
      "(n =", row$size,
      ", min cos sim =", row$min_cos_sim, ")\n\n")
  cat("**Medoid:**", row$medoid, "\n\n")

  # Plot medoid
  print(plot_89(medoid_sig, plot_title = paste0(
    "Cluster ", row$cluster_id, " medoid: ", row$medoid
  )))

  # If cluster is small enough, show all members overlaid
  members <- strsplit(row$members, "; ")[[1]]
  if (length(members) > 1) {
    member_mat <- combined[, members, drop = FALSE]
    # Plot each member as light gray, medoid in black
    cat("\n\nAll", length(members), "members overlaid:\n\n")
    # Simple matrix plot showing variation
    matplot(
      member_mat, type = "l", lty = 1,
      col = adjustcolor("gray50", alpha.f = 0.4),
      xlab = "Mutation type (index)", ylab = "Proportion",
      main = paste0("Cluster ", row$cluster_id, " — all members")
    )
    lines(combined[, row$medoid], col = "red", lwd = 2)
    legend("topright", legend = c("Members", "Medoid"),
           col = c("gray50", "red"), lwd = c(1, 2), cex = 0.8)
  }

  # Find and plot top 3 best-matching spectra from catalogs
  cos_sims <- apply(all_catalogs, 2, function(x) {
    lsa::cosine(medoid_sig[, 1], x)
  })
  top3_idx <- order(cos_sims, decreasing = TRUE)[1:min(3, length(cos_sims))]

  for (rank in seq_along(top3_idx)) {
    idx <- top3_idx[rank]
    spec_name <- names(cos_sims)[idx]
    spec_cos  <- cos_sims[idx]

    cat("\n\n**Best-matching spectrum", rank, ":**", spec_name,
        "(cosine similarity =", round(spec_cos, 4), ")\n\n")

    spec <- all_catalogs[, spec_name, drop = FALSE]
    print(plot_89(spec, plot_title = paste0(
      "Best-matching spectrum ", rank, ": ", spec_name,
      " (cos sim = ", round(spec_cos, 4), ")"
    )))
  }
}
```

## Save outputs

```{r save-outputs}
out_dir <- here::here("Manuscript_data", "Mo_CAP9_analysis", "clustering_results")
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

prefix <- "CAP9_Koh89_clustered"

# Summary table
write.table(
  summary_df,
  file = file.path(out_dir, paste0(prefix, "_summary.tsv")),
  sep = "\t", row.names = FALSE, quote = FALSE
)

# Medoid signatures matrix
medoid_mat <- combined[, names(medoid_names), drop = FALSE]
write.table(
  medoid_mat,
  file = file.path(out_dir, paste0(prefix, "_medoid_signatures.tsv")),
  sep = "\t", quote = FALSE
)

# All cluster assignments
assignments <- data.frame(
  signature = names(clusters),
  cluster_id = clusters,
  retained = clusters %in% retained_df$cluster_id,
  stringsAsFactors = FALSE
)
write.table(
  assignments,
  file = file.path(out_dir, paste0(prefix, "_all_cluster_assignments.tsv")),
  sep = "\t", row.names = FALSE, quote = FALSE
)

cat("Outputs saved to:", out_dir, "\n")
```

## Dropped clusters

```{r dropped}
if (nrow(dropped_df) > 0) {
  cat("Clusters dropped (size <", min_cluster_size, "):\n\n")
  DT::datatable(
    dropped_df[, c("cluster_id", "size", "medoid", "members")],
    rownames = FALSE,
    options = list(pageLength = 25, scrollX = TRUE),
    caption = "Dropped clusters"
  )
} else {
  cat("No clusters were dropped.\n")
}
```

::: {.callout-note collapse="true"}
## Session Info

```{r session-info}
#| echo: false
#| comment: ""
sessioninfo::session_info(info = "all")
```
:::
